### 2021-08-22 doc/thesis/chapters/results.tex
res: interfaces and sim time

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Lines of Code}

@@

This is the main interface for the data input and -output of the DSP chain.
In HLS it can be automatically inferred by adding pragma directives to the respective function parameter variables.
Also the testbench code is implemented like a regular function call.
Therefore, the actual interface logic is completely hidden from the developer.
In VHDL on the other hand, this logic has to be implemented manually.
Also the testbench has to be hooked up with the correct signals and have a verificator logic, that supports the functional verification of an AXI stream interface.
Otherwise, this logic would have to be implemented manually as well.
In the time comparison, the larger amount of manual implementation in VHDL is clearly visible.
It is to note that this comparison does not compare functionally equivalent implementations, because the actually implemented logic in VHDL does not use the ready-flag correctly, as explained in the previous chapters.
According to this fact, a further increase in implementation hours should actually be considered for VHDL.
However, also the HLS implementation has its particular characteristics that need to be understood for a fully functional implementation.
Following the respective user guides and implementation examples provides a good starting point for that.

@@

This interface is used to control IP settings and to read status information.
Similar to the previous AXI Stream interface, also this AXI4-Lite memory-mapped interface is automatically generated by HLS, while this has to be done manually in VHDL.
To create this process more efficiently and extendable, a Python script is used, which can also automatically generate VHDL code for the registers' memory map.
This turns out to be a multiplicator in productivity for the implementation of registers.
The register generator script was previously used in projects and is therefor already well-known and tested before it is used in this project.
However, the VHDL code only represents one part of the interface -- the second part is the firmware driver that is used in the application software.
Even though the register generator script also generates a \cplusplus\ header file that includes several address offset definitions of the registers, the actual driver functions need to be implemented manually.
Several files need to be integrated into a pre-defined folder structure, in order to enable a correct integration of the IP in a Vivado block diagram.\\

@@

Once the AXI4-Lite memory-mapped interface's main functionality is implemented, further registers can be added.
In the specific implementation for this project, registers for LED control, the DSP mode, as well as a build information register are added.
The latter holds information about the build date and -time of the IPs, and is particularly time-consuming because of the implementation of the respective drivers to convert the date format in both variants.
In contrast, the addition of the LED control and DSP mode registers is simple and fast, because of the auto-generated code in HLS and VHDL with usage of the register engine.
The single bits in the LED control register directly represent the LEDs' values on the hardware.
The DSP mode register can select the regular FM radio mode and a pass-through mode, which simply forwards the streaming data without any processing.
Summing up, the addition of registers to the design can be time-consuming in both variants, depending on the required driver functionality.
However, the general tendency is that VHDL requires more time, because of this driver issue.

@@

For the \cplusplus\ version, a code example for an FIR filter, which is provided by Xilinx HLS, is taken and modified for the present purpose.
The VHDL variant is taken from an older DSP project, which was previously developed during a university course.
However, the integration and testbench code is developed for this project, specifically.
Here, the HLS implementation took significantly longer, because of the limited knowledge regarding the usage of an AXI stream interface.
It is the first time in the project, that this type of interface is applied in the IP design and used in the testbench.
The VHDL implementation was straightforward in comparison.
The testbenches compare the FIR filters' output with the Matlab model's pilot tone output.
Both variants match the output results of the Matlab model successfully.

@@

The implementation of the sample decimation functionality differs slightly between the two variants, HLS and VHDL.
This is explained in Section~\ref{sec:impl:hls:sample_rate_reduction}.
The implementation time of the HLS version is severely larger than the VHDL version.
This is mainly due to the limited knowledge of the functionality of the AXI stream interface at this stage in the project's implementation, just as previously mentioned with the FIR filter.
The initial version was implemented like the VHDL version -- a counter that only triggers the subsequent DSP blocks upon a certain number of samples was received.
However, this led to an endlessly blocking AXI stream in the testbench execution.
The reason was, that the read-function can only read a sample when the respective stream contains a sample, and that it is a blocking function.
Because the sample decimator did not write a sample every time, the subsequent read-function blocked indefinitely.
A thorough investigation in several user guides, in combination with step-by-step debugging, the issue was found and led to the current, working implementation, which processes a number of samples, before it forwards one sample.
The process of finding this issue explains the largely varying implementation times, as listed in Table~\ref{tab:implementation_time_logging}.

@@

The implementation of the receiver structure requires the combination of several previously implemented blocks, such as the FIR filter and any interfaces.
This is done according to the Matlab model, to resemble its functionality.
The VHDL version is significantly more time consuming in comparison to HLS.
However, it is to note at this point that the VHDL version is implemented first, and thus some issues are already solved by the time of the HLS implementation.
Nevertheless, the VHDL implementation takes longer, as it requires more detailed care in the signal level.
This specifically includes the clock-alignment of any signals that need to be used in a combination.
For example, if two parallel FIR filters produce output signals that are to be added, they need to be aligned so that they are valid at the time of the summation.
This often requires analysis and debugging in a signal wave form.
In HLS this is not required, since the validity of a signal is determined internally by the HLS compiler, according to the usage of the respective data variables.
Another factor that adds up to the implementation time is the number of lines of code, as presented in Table~\ref{tab:lines_of_code}.
The VHDL IP design requires 3000 lines, while the HLS IP design only takes 920 for the same functionality.
Again, this is a subjective measure -- several factors such as code style, or the developer's experience play a role -- but still shows a good representation of the implementation effort for this project.\\

In the final summation over several implementation times of all the listed features, the two variants HLS and VHDL only differ by 10 hours, which is around 15\%.
However, as the previous paragraphs explain, prior knowledge regarding HLS was limited and had to be gathered during the ongoing implementation.
Nevertheless, even consiering this fact, the HLS implementation is faster than the VHDL equivalent.
Thus, with an equal amount of prior knowledge and experience, it is assumed that an HLS implementation can lead to a successfully functioning result significantly faster.

@@

Figure~\ref{fig:impl_time_analysis} shows this direct comparison of processed samples per second for the implementation variants.
The measurements always include the testbench setup-phase, and the run-phase.
The setup-phase mainly consists of loading the recorded data and its preparation, so that it can be fed into the DUT.
The run-phase then actually applies this test data to the DUT and stores the respective output results.
All time measurements are taken without the result analysis, since this is the exact same, shared code between all variants and thus takes the same amount of time everywhere.
Several measurements are done multiple times, to achieve an average result, that is independent of the current processor load of the host PC.\\

The diagram shows an immense difference of simulation speed between the variants.
It is important to note that the y-axis has a logarithmic scale, as a method to enable the presentation of the values in multiple orders of magnitude.
VHDL only achieves around 20 samples per second.
This is by far the slowest method, which is expected because of the way of how VHDL is simulated.
This can not be explained in detail here, but one main feature of a VHDL simulator is its capability to simulate parallel processes, as they occur in actual parallel hardware.
The VHDL simulator therefor needs to evaluate several concurrent processes and statements, whenever any of the interacting signals change their value, and this is time-consuming.
In the current case, there is an additional interface between the VHDL simulator and the Python testbench framework, which likely additionally slows down the simulation.
The HLS testbench in comparison achieves around 50000 samples per second, which is a speed-up factor of 2500.
This is specifically talking about the HLS C/\cplusplus-simulation.
The main reason for the speed-up therefor is, that in this type of simulation no timing behaviour is simulated -- any processes are treated as C/\cplusplus\ functions that runs like a regular executable on the host CPU.
Also, the HLS compiler is using native \cplusplus\ data types to represent the signals, where possible, which results in the faster execution of the code.
Finally, the simulation in Matlab achieves a further speed-up by a factor of 70.
Matlab is specifically designed for applications which require matrix calculations and utilizes highly optimized algorithms in its underlying functions therefor.
Thus, Matlab runs very efficiently and achieves the highest simulation speed of around 3 million samples per second.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Lines of Code}
\label{sec:results:lines-of-code}

@@

This is the main interface for the data input and -output of the DSP chain.
In HLS, it can be automatically inferred by adding pragma directives to the respective function parameter variables, and the testbench code is implemented like a regular function call.
Therefore, the actual interface logic is completely hidden from the developer.
In VHDL on the other hand, this logic has to be implemented manually.
Also the testbench has to be hooked up with the correct signals and have a verificator logic, that supports the functional verification of an AXI stream interface.
Otherwise, this logic would have to be implemented manually as well.
In the time comparison, the larger amount of manual implementation in VHDL is clearly visible.
It is to note, that this comparison does not compare functionally equivalent implementations though, because the actually implemented logic in VHDL does not use the ready-flag correctly, as explained in the previous chapters.
According to this fact, an even further increase in implementation hours should actually be considered for VHDL.
However, also the HLS implementation has its particular characteristics that need to be understood for a fully functional implementation, which requires time in research.
Following the respective user guides and implementation examples provides a good starting point for that.

@@

This interface is used to control IP settings and to read status information.
Similar to the previous AXI Stream interface, also this AXI4-Lite memory-mapped interface is automatically generated by HLS, while this has to be done manually in VHDL.
To create this process more efficiently and extendable, a Python script -- the so-called Register Engine -- is used, which can also automatically generate VHDL code for the registers' memory map.
This turns out to be a multiplicator in productivity for the implementation of registers.
The register generator script was already used in previous projects and is therefore well-known and tested, before it is used in this project.
However, the VHDL code only represents one part of the interface -- the second part is the firmware driver that is used in the application software.
Even though the register generator script already generates a \cplusplus\ header file, which includes several address offset definitions of the registers, the actual driver functions need to be implemented manually.
Furthermore, several files need to be integrated into a pre-defined folder structure that Vivado understands, in order to enable a correct integration of the IP in a Vivado block diagram.\\

@@

Once the AXI4-Lite memory-mapped interface's main functionality is implemented, further registers can be added.
In the specific implementation for this project, registers for LED control, the DSP mode, as well as a build information register are added.
The latter holds information about the build date and -time of the IPs, and is particularly time-consuming because of the implementation of the respective drivers to convert the date format in both variants.
In contrast, the addition of the LED control and DSP mode registers is simple and fast, because of the auto-generated code in HLS and VHDL, with usage of the register engine.
The single bits in the LED control register directly represent the LEDs' values on the hardware.
The DSP mode register can select between the regular FM radio mode and a pass-through mode, which simply forwards the streaming data without any processing.\\

Summing up, the addition of registers to the design can be time-consuming in both variants, depending on the required driver functionality.
However, the general tendency is, that VHDL requires more time, because of this driver issue.

@@

For the HLS version, a \cplusplus\ code example of an FIR filter, which is provided by Xilinx HLS, is taken and modified for the present purpose.
The VHDL variant is taken from an older DSP project, which was previously developed during a university course.
However, the integration- and testbench code is developed for this project, specifically.
Here, the HLS implementation took significantly longer, because of the limited knowledge regarding the usage of an AXI stream interface.
It is the first time in the project, that this type of interface is applied in the IP design and used in the testbench.
The VHDL implementation is straightforward, in comparison.
The testbenches compare the FIR filters' output with the Matlab model's pilot tone output.
Both variants match the output results of the Matlab model successfully.

@@

The implementation of the sample decimation functionality differs slightly between the two variants, HLS and VHDL.
This is explained in Section~\ref{sec:impl:hls:sample_rate_reduction}.
The implementation time of the HLS version is severely larger than the VHDL version.
This is mainly due to the limited knowledge of the functionality of the AXI stream interface at this stage in the project's implementation, just as previously mentioned with the FIR filter.
The initial version was implemented like the VHDL version -- a counter that only triggers the subsequent DSP blocks upon a certain number of samples is received.
However, this led to an endlessly blocking AXI stream in the testbench execution.
The reason was, that the read-function can only read a sample when the respective stream contains a sample, and that it is a blocking function.
Because the sample decimator did not write a sample every time, the subsequent read-function blocked indefinitely.
By a thorough investigation in several user guides, in combination with step-by-step debugging, the issue was found and led to the current, working implementation, which processes a number of samples, before it forwards one sample.
The process of finding this issue explains the largely varying implementation times, as listed in Table~\ref{tab:implementation_time_logging}.

@@

The implementation of the receiver structure requires the combination of several previously implemented blocks, such as the FIR filter instances and any interfaces.
This is done according to the Matlab model, to resemble its functionality.
The VHDL version is significantly more time consuming in comparison to HLS.
However, it is to note at this point, that the VHDL version is implemented first, and thus some issues are already solved by the time of the HLS implementation.
Nevertheless, the VHDL implementation takes longer, as it requires more detailed care in the signal level.
This specifically includes the clock-alignment of any signals that need to be used in a combination.
For example, if two parallel FIR filters, which have a different filter order and thus, a different group delay, produce output signals that are to be added, they need to be aligned, so that they are valid at the time of the summation.
This often requires analysis and debugging in a signal wave form.
In HLS this is not required, since the validity of a signal is determined internally by the HLS compiler, according to the usage of the respective data variables.
Another factor that adds up to the implementation time, is the number of lines of code, as presented in Section~\ref{sec:results:lines-of-code}.
The VHDL IP design requires 3000 lines, while the HLS IP design only takes 920 for the same functionality.
Again, this is a subjective measure -- several factors such as code style, or the developer's experience play a role -- but still shows a good representation of the implementation effort for this project.\\

In the final summation over several implementation times of all the listed features, the two variants HLS and VHDL only differ by 10 hours, which is about 15\%.
However, as the previous paragraphs explain, prior knowledge regarding HLS was limited and had to be gathered during the ongoing implementation.
Nevertheless, even considering this fact, the HLS implementation is faster than the VHDL equivalent.
Thus, with an equal amount of prior knowledge and experience, it is assumed that an HLS implementation can lead to a successfully functioning result significantly faster.

@@

Figure~\ref{fig:impl_time_analysis} shows this direct comparison of processed samples per second for the implementation variants.
The measurements always include the testbench setup-phase, and the run-phase.
The setup-phase mainly consists of loading the recorded data and its preparation, so that it can be fed into the DUT.
The run-phase then actually applies this test data to the DUT and stores the respective output results.
All time measurements are taken without the result analysis, since this is the exact same, shared code between all variants and thus, takes the same amount of time everywhere.
Several measurements are done multiple times, to achieve an average result, that is independent of the current processor load of the host PC.\\

The diagram shows an immense difference of simulation speed between the variants.
It is important to note, that the y-axis has a logarithmic scale, as a method to enable the presentation of the values in multiple orders of magnitude in a single diagram.
VHDL only achieves around 20 samples per second.
This is by far the slowest method, which is expected, because of the way of how VHDL is simulated.
This can not be explained in detail here, but one main feature of a VHDL simulator is its capability to simulate parallel processes, as they occur in actual parallel hardware.
The VHDL simulator therefor needs to evaluate several concurrent processes and statements, whenever any of the interacting signals change their value, and this is time-consuming.
In the current case, there is also an additional interface between the VHDL simulator and the Python testbench framework, which likely slows down the simulation even further.
The HLS testbench in comparison achieves around 50.000 samples per second, which is a speed-up factor of 2.500.
This is specifically talking about the HLS C/\cplusplus\ simulation.
The main reason for the speed-up is, that this type of simulation does not simulate any timing behaviour -- any processes are treated as C/\cplusplus\ functions, which run like a regular executable on the host CPU.
Also, the HLS compiler is, where possible, using native \cplusplus\ data types to represent the signals, which results in the faster execution of the code.
Finally, the simulation in Matlab achieves a further speed-up by a factor of 70.
Matlab is specifically designed for applications, which require matrix calculations and utilizes highly optimized algorithms in its underlying functions, therefor.
Consequently, Matlab runs very efficiently and achieves the highest simulation speed of around 3~million samples per second.
